---
title: "fredmr - loading and transforming FRED-MD"
author: "Philipp Hauber"
date: "28 10 2019"
output: html_document
---

## Load csv file

```{r}
# clear workspace
rm(list = ls())

# set working directory
#setwd("C:/Users/Philipp/Documents/GitHub/fredmdr")
setwd("C:/Users/Hauber/Desktop/fredmdr")

# load in data
temp <- read.csv("2019-09.csv")
```

## Extract series, transformations, names and dates

```{r}
# extract trafos
trafos <- temp[1, 2 : dim(temp)[2]]
xmat <- temp[2 : dim(temp)[1], 2 : dim(temp)[2]]

# convert first column to date format and move to rownames
rownames(xmat) <- seq(as.Date("1959-01-01"), by = "months", length.out = dim(xmat)[1])

# remove temp
rm(temp)
```

Variable names are stored as column names of xmat

## Transform series

First, define a function that transforms the series according to its code: 

```{r}
f_transform <- function(x, trafos){
  
  # back out dimensions
  Nt <- dim(x)[1] ;  Nn <- dim(x)[2]
  # check that number of columns equals length of trafos
  if (Nn != length(trafos)){
    print("dimensions of x and trafos do not match. Abort!")
    break
  }
  
  # output matrix (carrying over row and col names)
  xtrafo <- x
  xtrafo[] <- NA
  
  # loop over Nn
  for (i in 1:Nn) {
    if (trafos[i] == 1) # no transformation
      xtrafo[, i] <- x[, i]
    
    else if (trafos[i] == 2) # first difference
      xtrafo[ 2 : Nt, i] <- diff(x[, i])
    
    else if (trafos[i] == 3) # second difference
      xtrafo[ 3 : Nt, i] <- diff(x[, i], differences = 2)
    
    else if (trafos[i] == 4) # logs
      xtrafo[, i] <- log(x[, i])
    
    else if (trafos[i] == 5) # log first differences
      xtrafo[ 2 : Nt, i] <- diff(log(x[, i]))
    
    else if (trafos[i] == 6) # log second differenes
      xtrafo[ 3 : Nt, i] <- diff(log(x[, i]), differences = 2)
    
    else if (trafos[i] == 7) # difference of percentage change
      xtrafo[ 3 : Nt, i] <- diff(x[2 : Nt, i] / x[1 : (Nt-1), i] - 1)
  } 
  return(xtrafo)
}
```

Now, transform columns of **xmat** calling **f_transform**. 

```{r}
xmat <- f_transform(xmat, trafos)
```

Remove the first two rows which are mostly NA due to transformations

```{r}
xmat <- xmat[3 : dim(xmat)[1], ]
```

## Remove outliers

function

```{r}
f_removeoutl <- function(x){
  ########################################################    
  # removes outliers from Nt x Nn matrix x, 
  # defined as any observation larger than 10 times 
  # the interquartile range (in absolute value)
  ########################################################    
    
  # back out dimensions of x    
  Nt <- dim(x)[1] ; Nn <- dim(x)[2]
  
  # initialize xoutl (carrying over row and column names)
  xoutl <- x
  
  # calculate median of columns of x
  medx <- matrix(apply(x, 2, median, na.rm = TRUE), nrow = Nt, ncol = Nn, byrow = TRUE)
  
  # calculate interquartile range of columns of x
  iqrx <- matrix(apply(x, 2, quantile, probs = c(0.75), na.rm = TRUE) - 
                 apply(x, 2, quantile, probs = c(0.25), na.rm = TRUE), 
                 nrow = Nt, ncol = Nn, byrow = TRUE)
  
  # determine outliers
  z <- abs(x-medx);
  outlier <- z > (10 * iqrx)
  
  # remove and count
  xoutl[outlier == TRUE] <- NA
  noutl <- apply(outlier, 2, sum, na.rm = TRUE)
  
  # return output
  return(list(xoutl = xoutl, noutl = noutl))
}
```

Call **f_removeoutl** and unlist results
```{r}
temp <- f_removeoutl(xmat)

# unlist!
xmat <- temp[["xoutl"]]
noutl <- temp[["noutl"]]
rm(temp)
```

## Compare with MATLAB

Load series transformed by MATLAB

```{r}
xmatlab <- read.csv("original MATLAB code/dataMATLAB.csv", header = FALSE)
```

Plot!

```{r}
ind_var <- which(names(xmat) == "PAYEMS")
diffRMATLAB <- xmat[, ind_var] - xmatlab[, ind_var]
plot(diffRMATLAB, 
     type = "l", xlab = "", ylab = "", 
     main = paste0("Difference between MATLAB and R: ", names(xmat)[ind_var]))
```

Virtually no differences in the log difference of non-farm employment!

## Standardize series

Function with default arguments such that mean is set to 0 and standard deviation to 1

```{r}
f_standardize <- function(x, demean = TRUE, unitsd = TRUE) {
  ########################################################    
  # standardizes Nt x Nn matrix x (NA are allowed) #######
  ########################################################  
  Nt <- dim(x)[1] ; Nn <- dim(x)[2] ; xstand <- x
  
  # claculate mean and sd of columns of x
  meanx <- matrix(apply(xstand, 2, mean, na.rm = TRUE), nrow = Nt, ncol = Nn, byrow = TRUE)
  sdx <- matrix(apply(xstand, 2, sd, na.rm = TRUE), nrow = Nt, ncol = Nn, byrow = TRUE)
  if (demean == TRUE & unitsd == TRUE) {
    xstand <- xstand - meanx
    xstand <- xstand / sdx
  } else if (demean == TRUE & unitsd == FALSE) {
    xstand <- xstand - meanx
  } else if (unitvar == TRUE & demean == FALSE) {
    xstand <- xstand / sdx
  }
  return(list(xstand = xstand, meanx = meanx[1, ], sdx = sdx[1, ]))
}
```

Call **f_standardize** and unlist results

```{r}
temp <- f_standardize(xmat)

# unlist
xmat <- temp[["xstand"]]
meanx <- temp[["meanx"]]
names(meanx) <- names(xmat)
sdx <- temp[["sdx"]]
names(sdx) <- names(xmat)
rm(temp)
```

## Re-insert dates as column and tidy up

```{r}
xmat <- cbind(rownames(xmat), data.frame(xmat, row.names=NULL))
names(xmat)[1] <- "dates"
library(tidyr)
fredmd <- gather(xmat, key = var, value = value, -dates)
```


## Save data frame

```{r}
save(fredmd, file = "fredmd201909.Rda")
```

